Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-06-19 manuscript/manuscript.tex
Fix tabel labeling
Change author order
Address Franz comments in the introduction

\author[1\authfn{1}]{Franz Waibl}
\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[2]{Vjay Molino}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[2]{Steven Ramsey}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system. Towards this, Jupyter notebooks and python packages are provided to simplify the analysis.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.

@@

Interaction with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent \cite{ben-naim-book}.
It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.

Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
It estimates the solvent density from an MD trajectory and computes localized entropic and enthalpic contributions on a grid around the solute molecule.
Different from most other methods, these contributions are therefore spatially resolved, improving interpretability and allowing better insights into the nature of solvation around the investigated solute.
The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.

@@

Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
As such, it has been deployed in systematically mapping the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.

@@

In this tutorial, we aim to give the reader an introduction into GIST and its application to calculate solution thermodynamics properties of interest. 
While previous publications \cite{Ramsey2016} and tutorials\footnote{See \url{http://ambermd.org/tutorials/advanced/tutorial25/}} gave insightful pointers how to setup and run GIST calculations, the post-processing was often treated less rigorously. While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
Furthermore, we aim to establish a template for the analysis and interpretation of GIST calculations.

@@

Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
Two major flavors of methods in this group are those based on the Ornstein--Zernike (OZ) equation \cite{Hansen2013-simple-liquids} and those based on inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is their reliance on approximating the entropy through an infinite expansion in terms of the density distribution. 
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.

@@

	\label{tab:ref_values}
	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
	\small
	\begin{tabularx}{\columnwidth}{@{}lrrr@{}}
		\toprule
		Water Model       & $\rho_0$ & $E_{ww, norm}$ & $N_\text{WAT}$ \\
        & $/\text{\AA}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
		\midrule
		TIP3P \cite{Jorgensen1983-tip3p}    & 0.03287 & -9.5398   & 31795 \\
		TIP3PFB \cite{Wang2014-tip3p-force-balance}  & 0.03321 & -11.7214  & 32502 \\
		TIP4P \cite{Jorgensen1985-tip4p}    & 0.03316 & -9.8583   & 31218 \\
       TIP4PFB \cite{Wang2014-tip3p-force-balance} & 0.03330 & -11.8724   & 31653 \\
		TIP4P/Ew \cite{Horn2004-tip4pew} & 0.03323 & -11.0541  & 31218 \\
		TIP5P \cite{Mahoney2000-tip5p}    & 0.03285 & -9.5989   & 31382 \\
		SPC/E \cite{Berendsen1987-spce}    & 0.03333 & -11.1334  & 31795 \\
       SPC/Eb \cite{Takemura2012-spceb}   & 0.03358 & -11.6966 & 31795 \\
		OPC \cite{Izadi2014-opc}      & 0.03330 & -12.2456 & 31842 \\
		OPC3 \cite{Izadi2016-opc3}     & 0.03323 & -11.6994 & 32251 \\
		\bottomrule
	\end{tabularx}
\end{table}

@@

The Amber manual provides reference densities and energies for several solvents.
Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_values}.
However, the exact energy values are different when using PME-GIST, and also depend on the box size.
For quantitative analyses, it is recommended to compute an exact reference energy.
The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per solvent molecule.

>>>>

\author[1\authfn{1}]{Valentin J Hoerschinger}
\author[1\authfn{1}]{Franz Waibl}
\author[2]{Vjay Molino}
\author[2]{Helmut Carter}
\author[1]{Monica L Fern{\'a}ndez-Quintero}
%\author[1,2]{Firstname Middlename Familyname}
%\author[2]{Firstname Initials Surname}
\author[2]{Steven Ramsey}
\author[4]{Daniel R Roe}
\author[1*]{Klaus R Liedl}
\author[3*]{Michael K Gilson}
\author[2*]{Tom Kurtzman}

@@

In this tutorial, we aim to guide the reader through the most common steps involved in a GIST analysis using the streptavidin-biotin complex as an demonstrative system. Towards this, Jupyter notebooks and Python packages are provided to simplify the analysis.
Furthermore, we discuss the theory of GIST with a focus on practical aspects.
We highlight potential pitfalls and show how to avoid technical difficulties.
For this tutorial, familiarity with molecular dynamics simulations as well as the AmberTools package \cite{Case2023-ambertools} is assumed.

@@

Interactions with water are of special importance as many biological processes occur in aqueous environment \cite{Privalov2017-water-review}.
The free energy of solvation is defined as the reversible work needed to insert a solute molecule into the solvent from gas phase \cite{ben-naim-book}.
It is directly linked to the solubility of gases and drives many real-world effects such as hydrophobicity, ligand binding, and protein folding.

Grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012} is a method to compute the free energy of solvation, with applications in the context of drug design and biophysical property description.
It computes the solvent density from an MD trajectory and estimates localized entropic and enthalpic contributions on a grid around the solute molecule.
Since these contributions are spatially resolved, GIST improves interpretability and offers more detailed insights into the solvation of the investigated solute. 
The entropy is computed from the density distribution of the solvent around the solute molecule, in a first-order approximation. A schematic of this is shown in Figure~\ref{fig:carbazole}. 
The calculation of higher-order entropy terms is computationally demanding but was realized in recent works \cite{Nguyen2016-gist-second-order, Waibl2021-gist-salt}.
Nevertheless, solvation free energies of small molecule solutes can be computed at high accuracy using linear scaling factors for the higher-order terms, such as $\Delta S = 0.6\, \Delta S^\text{1st-order}$ for water \cite{Chen2021,Waibl2022-gist-solvents}.

@@

Since its initial development, GIST has been used to describe the thermodynamics of water displacement in protein binding pockets \cite{Nguyen2016-gist-second-order,Ramsey2016,Balius2017-gist-ligand-discovery,Hufner-Wulsdorf2020-gist-drug-design}.
It has been used for a large-scale investigation of the hydration properties of proteins found in SARS-CoV-2 \cite{Olson2020-covid-gist}.
Furthermore, recent works used GIST to describe biophysical properties of serine proteases \cite{Kraml2019-gigist}, macrocycles \cite{Kamenik2020-gist-macrocycles}, or antibody fragements \cite{Waibl2021-gist-antibodies}.

@@

In this tutorial, we aim to give the reader an introduction to GIST and its application to calculate solution thermodynamics properties of interest. 
While previous publications \cite{Ramsey2016} and tutorials \cite{amber_tut_gist} treated how to setup and run GIST calculations in detail, the post-processing was often treated less rigorously. While building on these tutorials, we provide an updated guide to calculating solvation properties using GIST to address recent improvements.
Furthermore, we aim to establish best practices for the analysis and interpretation of GIST calculations.

@@

Methods derived from statistical mechanics bridge this gap by providing three-dimensional resolution.
These methods most commonly are based on either the Ornstein–Zernike (OZ) equation, or \cite{Hansen2013-simple-liquids} inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998}.
The most important example of OZ-based methods is the reference interaction site model (RISM) \cite{Chandler1972-rism} as well as its extension to three dimensions (3D-RISM) \cite{Kovalenko1998-3drism}.
Those methods compute the atomic solvent distributions by self-consistently solving the OZ integral equation to predict solvation thermodynamics.

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
%TODO add MIE citation and context
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.

@@

	\caption{Reference number density $\rho_0$ and mean solvent-solvent energy $E_{ww, norm}$ for various water models from NPT simulations of N$_{WAT}$ water solvent molecules at \SI{300}{\kelvin} and 1bar.}\label{tab:ref_densities}
	\small
	\begin{tabularx}{\columnwidth}{@{}lrrr@{}}
		\toprule
		Water Model       & $\rho_0$ & $E_{ww, norm}$ & $N_\text{WAT}$ \\
        & $/\text{\AA}^{-3}$ & $/\text{kcal}\cdot \text{mol}^{-1}$ & \\
		\midrule
		TIP3P \cite{Jorgensen1983-tip3p}    & 0.03287 & -9.5398   & 31795 \\
		TIP3PFB \cite{Wang2014-tip3p-force-balance}  & 0.03321 & -11.7214  & 32502 \\
		TIP4P \cite{Jorgensen1985-tip4p}    & 0.03316 & -9.8583   & 31218 \\
       TIP4PFB \cite{Wang2014-tip3p-force-balance} & 0.03330 & -11.8724   & 31653 \\
		TIP4P/Ew \cite{Horn2004-tip4pew} & 0.03323 & -11.0541  & 31218 \\
		TIP5P \cite{Mahoney2000-tip5p}    & 0.03285 & -9.5989   & 31382 \\
		SPC/E \cite{Berendsen1987-spce}    & 0.03333 & -11.1334  & 31795 \\
       SPC/Eb \cite{Takemura2012-spceb}   & 0.03358 & -11.6966 & 31795 \\
		OPC \cite{Izadi2014-opc}      & 0.03330 & -12.2456 & 31842 \\
		OPC3 \cite{Izadi2016-opc3}     & 0.03323 & -11.6994 & 32251 \\
		\bottomrule
	\end{tabularx}
\end{table}

@@

The Amber manual provides reference densities and energies for several solvents.
Furthermore, a large variety of reference values calculated with GPU-GIST are listed in table \ref{tab:ref_densities}. 
However, the exact energy values are different when using PME-GIST, and also depend on the box size.
For quantitative analyses, it is recommended to compute an exact reference energy.
The most accurate method to compute a reference energy is to run a GIST calculation of the pure solvent using the same energy method (PME/GPU), a similar box size, at the same temperature and pressure and then compute the average solvent-solvent energy per solvent molecule.
